Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2020-12-30 jss/EpiCompare-paper/EpiCompare-paper.Rmd
introduction update (message improvement)

@@

The recent (and currently on-going) COVID-19 global pandemic has galvanized public interest in understanding more about infectious disease modeling and has highlighted the usefulness of research in the area of infectious disease epidemiology. Infectious disease models typically attempt to do one or more of the following: 1) predict the spread of current and future epidemics \citep[e.g. flue prediction][]{Biggerstaff2016}, 2) analyze past and current epidemics to increase scientific knowledge \citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or project epidemic scenarios under pre-specified parameters \citep[e.g. ...][]{}. The COVID-19 pandemic highlights how all three goals are important both separately and taken as a whole.  Infectious diseases inflict enormous burdens on the world: millions of lives lost and trillions of dollars spent yearly.  Correctly analyzing and addressing these issues aids in prevention and mitigation of future outbreaks. \shannon{really like this paragraph}

@@

At the same time, analysis packages often only address a portion of the analysis pipeline. Modeling tools often don't provide easy ways to compare and assess their models on new data. Moreover, exploring and modeling epidemics require transforming and \textit{tidying} data in different ways. To fill these gaps, we present our our \proglang{R} package \pkg{EpiCompare}. Our package's primary focus is to aid and advance research in the area of comparison and assessment of epidemic \& epidemiological models. In Figure \ref{fig:pipeline}, we illustrate the data analysis pipeline of infectious diseases  as 1) data pre-processing, 2) exploratory data analysis (EDA), 3) modeling and simulating, 4) post-processing, and 5) comparison and assessment; where each previous part of the pipeline influences the next. \pkg{EpiCompare} provides tools to aids practitioners in all areas of this pipeline.

@@

This paper is broken up into the following sections, section \ref{sec:time-invariant} motivates and showcases tools of time-invariant analysis,  section \ref{sec:overview} presents an outline of how \pkg{EpiCompare} aids a practitioner in every step of the pipeline and section \ref{sec:tour} provides a thorough demonstrating of the tools through a detailed example of a full data analysis pipeline.

@@

$R_0$, the initial reproduction number, has been called the ``most important quantity in epdiemiology'' \citep[][]{heesterbeek2003}. By definition, $R_0$ is the number of expected secondary infections when a primary infection is introduced to a susceptible population. To see the importance of $R_0$, one need only read the newspaper \citep{Fisher2020} or look at the length of the table of estimated $R_0$ quantities for COVID-19 \citep{Aronson2020}. $R_0$ is also, maybe, the most famous \textit{time-invariant} numerical summary of an epidemic, which allows epidemics to be compared to one another in both different time and geographic scales.

@@

Beyond ternary plots, and the associated SIR models, we can extend the time-invariant approach described in Section \ref{sec:ternary} to higher dimensions (i.e. models with more states). The constraints in 3d that are met with the SIR model (that is $\sum_{i=1}^3 (\text{number in state}(i)) = N(t)$) actually represents a spaces of 3d simplexes, and the ternary plot specifically represents these after scaling to examine such values as proportions (ternary plots are known as a 3d unit simplex due to it's scaling). This same scaling for larger models (i.e. with more states) can be done onto different simplexes. In this package we present tools to help compare models (mostly through similations), and this tool compares these objects after projecting them into a one-dimension less space through the simplexical structure of the data.

>>>>

devtools::load_all("../../")

@@

The recent (and on-going) COVID-19 global pandemic has galvanized public interest in understanding more about infectious disease modeling and has highlighted the usefulness of research in the area of infectious disease epidemiology. Infectious diseases inflict enormous burdens on the world: millions of lives lost and trillions of dollars spent yearly. Infectious disease models typically attempt to do one or more of the following: 1) predict the spread of current and future epidemics \citep[e.g. flue prediction][]{Biggerstaff2016}, 2) analyze past and current epidemics to increase scientific knowledge \citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or project epidemic scenarios under pre-specified parameters \citep[e.g. ...][]{}^[\@Shannon: CITE]. At the same time, descriptive statistics and visualizations (from universities like John Hopkins \citep{}^[\@Shannon: CITE], many branches and levels of government, and new organizations \citep{}^[\@Shannon: CITE]) are an important first step of the process.

With the many visualization and exploratory tools, models and modeling paradigms, and reviews and comparisons in the literature and through MIDAS (Models of Infectious Disease Agent Study) Control Center comparison guides \citep{}^[\@Shannon: CITE], this field has a lot of devices to aid an individual practitioner decide the correct approach. At the same time, analysis packages often only address a portion of the analysis pipeline. Modeling tools often don't provide easy ways to compare and assess their models on new data. Moreover, exploring and modeling epidemics require transforming and \textit{tidying} data in different ways. To fill these gaps, we present our \proglang{R} package \pkg{EpiCompare}. Our package's primary focus is to aid and advance research in the area of comparison and assessment of epidemic \& epidemiological models. In Figure \ref{fig:pipeline}, we illustrate the data analysis pipeline of infectious diseases  as 1) data pre-processing, 2) exploratory data analysis (EDA), 3) modeling and simulating, 4) post-processing, and 5) comparison and assessment; where each previous part of the pipeline influences the next. \pkg{EpiCompare} provides tools to aids practitioners in all areas of this pipeline.

<!-- # old draft
The recent (and currently on-going) COVID-19 global pandemic has galvanized public interest in understanding more about infectious disease modeling and has highlighted the usefulness of research in the area of infectious disease epidemiology. Infectious disease models typically attempt to do one or more of the following: 1) predict the spread of current and future epidemics \citep[e.g. flue prediction][]{Biggerstaff2016}, 2) analyze past and current epidemics to increase scientific knowledge \citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or project epidemic scenarios under pre-specified parameters \citep[e.g. ...][]{}. The COVID-19 pandemic highlights how all three goals are important both separately and taken as a whole.  Infectious diseases inflict enormous burdens on the world: millions of lives lost and trillions of dollars spent yearly.  Correctly analyzing and addressing these issues aids in prevention and mitigation of future outbreaks. \shannon{really like this paragraph}

@@

At the same time, analysis packages often only address a portion of the analysis pipeline. Modeling tools often don't provide easy ways to compare and assess their models on new data. Moreover, exploring and modeling epidemics require transforming and \textit{tidying} data in different ways. To fill these gaps, we present our our \proglang{R} package \pkg{EpiCompare}. Our package's primary focus is to aid and advance research in the area of comparison and assessment of epidemic \& epidemiological models. In Figure \ref{fig:pipeline}, we illustrate the data analysis pipeline of infectious diseases  as 1) data pre-processing, 2) exploratory data analysis (EDA), 3) modeling and simulating, 4) post-processing, and 5) comparison and assessment; where each previous part of the pipeline influences the next. \pkg{EpiCompare} provides tools to aids practitioners in all areas of this pipeline. 
-->

@@

This paper is broken up into the following sections, section \ref{sec:time-invariant} motivates and showcases tools of time-invariant analysis,  section \ref{sec:overview} prxesents an outline of how \pkg{EpiCompare} aids a practitioner in every step of the pipeline and section \ref{sec:tour} provides a thorough demonstrating of the tools through a detailed example of a full data analysis pipeline.

@@

$R_0$, the initial reproduction number, has been called the "most important quantity in epdiemiology" \citep[][]{heesterbeek2003}. By definition, $R_0$ is the number of expected secondary infections when a primary infection is introduced to a susceptible population. To see the importance of $R_0$, one need only read the newspaper \citep{Fisher2020} or look at the length of the table of estimated $R_0$ quantities for COVID-19 \citep{Aronson2020}. $R_0$ is also, maybe, the most famous \textit{time-invariant} numerical summary of an epidemic, which allows epidemics to be compared to one another in both different time and geographic scales.

@@

Beyond ternary plots, and the associated SIR models, we can extend the time-invariant approach described in Section \ref{sec:r0} to higher dimensions (i.e. models with more states). The constraints in 3d that are met with the SIR model (that is $\sum_{i=1}^3 (\text{number in state}(i)) = N(t)$) actually represents a spaces of 3d simplexes, and the ternary plot specifically represents these after scaling to examine such values as proportions (ternary plots are known as a 3d unit simplex due to it's scaling). This same scaling for larger models (i.e. with more states) can be done onto different simplexes. In this package we present tools to help compare models (mostly through similations), and this tool compares these objects after projecting them into a one-dimension less space through the simplexical structure of the data.

### 2020-12-30 jss/EpiCompare-paper/EpiCompare-paper.tex
introduction update (message improvement)

The recent (and currently on-going) COVID-19 global pandemic has
galvanized public interest in understanding more about infectious
disease modeling and has highlighted the usefulness of research in the
area of infectious disease epidemiology. Infectious disease models
typically attempt to do one or more of the following: 1) predict the
spread of current and future epidemics
\citep[e.g. flue prediction][]{Biggerstaff2016}, 2) analyze past and
current epidemics to increase scientific knowledge
\citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or
project epidemic scenarios under pre-specified parameters
\citep[e.g. ...][]{}. The COVID-19 pandemic highlights how all three
goals are important both separately and taken as a whole. Infectious
diseases inflict enormous burdens on the world: millions of lives lost
and trillions of dollars spent yearly. Correctly analyzing and
addressing these issues aids in prevention and mitigation of future
outbreaks. \textcolor{orange}{really like this paragraph}

The current epidemic of COVID-19 also highlights that infectious disease
models are only one piece of the overall analysis pipeline. University
based resources like John Hopkins' and government numerical dashboards
(across all levels of government) during the COVID-19 epidemic remind us
that descriptive statistics and visualization can be a important first
step in the process (multiple \cite{}?).\}
\textcolor{orange}{add comment about NYT or something} Still, rightly
so, a large amount of theoretical work goes into modeling epidemics,
with different models focusing at the individual / agent level, network
structure or just aggregate flows (review paper \cite{}?). All placing
individuals / proportions of the populations into different states
(e.g.~suspectible, exposed, infected, recovered, etc.). With all these
models, review and comparison papers in the literature and through MIDAS
(Models of Infectious Disease Agent Study) Control Center helps the
individual practitioner decide the correct approach.
\textcolor{orange}{along with expertise from healthcare professionals...}

At the same time, analysis packages often only address a portion of the
analysis pipeline. Modeling tools often don't provide easy ways to
compare and assess their models on new data. Moreover, exploring and
modeling epidemics require transforming and \textit{tidying} data in
different ways. To fill these gaps, we present our our \proglang{R}
package \pkg{EpiCompare}. Our package's primary focus is to aid and
advance research in the area of comparison and assessment of epidemic \&
epidemiological models. In Figure \ref{fig:pipeline}, we illustrate the
data analysis pipeline of infectious diseases as 1) data pre-processing,
2) exploratory data analysis (EDA), 3) modeling and simulating, 4)
post-processing, and 5) comparison and assessment; where each previous
part of the pipeline influences the next. \pkg{EpiCompare} provides
tools to aids practitioners in all areas of this pipeline.

@@

\(R_0\), the initial reproduction number, has been called the ``most
important quantity in epdiemiology'\,' \citep[][]{heesterbeek2003}. By
definition, \(R_0\) is the number of expected secondary infections when
a primary infection is introduced to a susceptible population. To see
the importance of \(R_0\), one need only read the newspaper
\citep{Fisher2020} or look at the length of the table of estimated
\(R_0\) quantities for COVID-19 \citep{Aronson2020}. \(R_0\) is also,
maybe, the most famous \textit{time-invariant} numerical summary of an
epidemic, which allows epidemics to be compared to one another in both
different time and geographic scales.

@@

Beyond ternary plots, and the associated SIR models, we can extend the
time-invariant approach described in Section \ref{sec:ternary} to higher
dimensions (i.e.~models with more states). The constraints in 3d that
are met with the SIR model (that is
\(\sum_{i=1}^3 (\text{number in state}(i)) = N(t)\)) actually represents
a spaces of 3d simplexes, and the ternary plot specifically represents
these after scaling to examine such values as proportions (ternary plots
are known as a 3d unit simplex due to it's scaling). This same scaling
for larger models (i.e.~with more states) can be done onto different
simplexes. In this package we present tools to help compare models
(mostly through similations), and this tool compares these objects after
projecting them into a one-dimension less space through the simplexical
structure of the data.

@@

\caption{\label{tab:hags-extreme}The extremeness of the true simulations based on comparing psuedo-density estimates between true vs simulated curves}
\centering
\begin{tabular}[t]{l>{\raggedleft\arraybackslash}p{6cm}>{\raggedleft\arraybackslash}p{6cm}}
\toprule
Type & simulations-based estimated psuedo-density & proportion of simulations with lower estimated psuedo-density\\
\midrule
Simple SIR & 0.0036733 & 0.00\\
EpiModel SIR & 0.0149686 & 0.02\\
\bottomrule
\end{tabular}
\end{table}

>>>>

The recent (and on-going) COVID-19 global pandemic has galvanized public
interest in understanding more about infectious disease modeling and has
highlighted the usefulness of research in the area of infectious disease
epidemiology. Infectious diseases inflict enormous burdens on the world:
millions of lives lost and trillions of dollars spent yearly. Infectious
disease models typically attempt to do one or more of the following: 1)
predict the spread of current and future epidemics
\citep[e.g. flue prediction][]{Biggerstaff2016}, 2) analyze past and
current epidemics to increase scientific knowledge
\citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or
project epidemic scenarios under pre-specified parameters
\citep[e.g. ...][]{}\footnote{@Shannon: CITE}. At the same time,
descriptive statistics and visualizations (from universities like John
Hopkins \citep{}\footnote{@Shannon: CITE}, many branches and levels of
government, and new organizations \citep{}\footnote{@Shannon: CITE}) are
an important first step of the process.

With the many visualization and exploratory tools, models and modeling
paradigms, and reviews and comparisons in the literature and through
MIDAS (Models of Infectious Disease Agent Study) Control Center
comparison guides \citep{}\footnote{@Shannon: CITE}, this field has a
lot of devices to aid an individual practitioner decide the correct
approach. At the same time, analysis packages often only address a
portion of the analysis pipeline. Modeling tools often don't provide
easy ways to compare and assess their models on new data. Moreover,
exploring and modeling epidemics require transforming and
\textit{tidying} data in different ways. To fill these gaps, we present
our \proglang{R} package \pkg{EpiCompare}. Our package's primary focus
is to aid and advance research in the area of comparison and assessment
of epidemic \& epidemiological models. In Figure \ref{fig:pipeline}, we
illustrate the data analysis pipeline of infectious diseases as 1) data
pre-processing, 2) exploratory data analysis (EDA), 3) modeling and
simulating, 4) post-processing, and 5) comparison and assessment; where
each previous part of the pipeline influences the next. \pkg{EpiCompare}
provides tools to aids practitioners in all areas of this pipeline.

@@

\(R_0\), the initial reproduction number, has been called the ``most
important quantity in epdiemiology'' \citep[][]{heesterbeek2003}. By
definition, \(R_0\) is the number of expected secondary infections when
a primary infection is introduced to a susceptible population. To see
the importance of \(R_0\), one need only read the newspaper
\citep{Fisher2020} or look at the length of the table of estimated
\(R_0\) quantities for COVID-19 \citep{Aronson2020}. \(R_0\) is also,
maybe, the most famous \textit{time-invariant} numerical summary of an
epidemic, which allows epidemics to be compared to one another in both
different time and geographic scales.

@@

Beyond ternary plots, and the associated SIR models, we can extend the
time-invariant approach described in Section \ref{sec:r0} to higher
dimensions (i.e.~models with more states). The constraints in 3d that
are met with the SIR model (that is
\(\sum_{i=1}^3 (\text{number in state}(i)) = N(t)\)) actually represents
a spaces of 3d simplexes, and the ternary plot specifically represents
these after scaling to examine such values as proportions (ternary plots
are known as a 3d unit simplex due to it's scaling). This same scaling
for larger models (i.e.~with more states) can be done onto different
simplexes. In this package we present tools to help compare models
(mostly through similations), and this tool compares these objects after
projecting them into a one-dimension less space through the simplexical
structure of the data.

@@

\caption{\label{tab:hags-extreme}The extremeness of the true simulations based on comparing psuedo-density estimates between true vs simulated curves}
\centering
\begin{tabular}[t]{l>{\raggedleft\arraybackslash}p{6cm}>{\raggedleft\arraybackslash}p{6cm}}
\toprule
Type & simulations-based estimated psuedo-density & proportion of simulations with lower estimated psuedo-density\\
\midrule
Simple SIR & 0.0036733 & 0\\
EpiModel SIR & 0.0028813 & 0\\
\bottomrule
\end{tabular}
\end{table}
